>

### 2019-04-05 paper/spinifex_paper.Rmd
paper edits through step 3

---
title: > 
  spinifex: manual control of dynamic linear projections of high-dimensional data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - School of Information Technology
    - Monash University, Melbourne, VIC 3800, Australia
    email:  Nicholas.Spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - School of Econometrics and Business Statistics
    - Monash University, Melbourne, VIC 3800, Australia
    email:  DiCook@monash.edu
abstract: >
  The class of dynamic linear projections that are collectively known as 'tours' provide a unique dynamic visualization of numeric multivariate data. Tours are particularly useful for understanding the structure held within multivariate data, and in association with techniques for dimension reduction, supervised, and unsupervised classification. The *R* package *tourr* offers a variety of path generators and geometric displays for conducting tours. This paper discusses an extension package, *spinifex*, that adds support for the path generation of manual tours and extends the display of tours to use with the contemporary animation packages, *plotly* and *gganimate*. Manual tours are used to explore the sensitivity of structure as the contributions of a manipulation variable are changed. This particularly useful after identifying a feature of interest.

  A recent paper {@wang_mapping_2018} visualizes the sensitivey of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolves more structrual information that is orthogonal to the original viewing plane.
keywords: manual tour, guided tour, grand tour, projection pursuit, high dimensional data, multivariate data, data visualization, statistical graphics, data science
output:
  bookdown::pdf_book:
    #base_format: rticles::rjournal_article #see sand box for hard code solution.
    fig_caption: yes
    includes:
      in_header: preamble.tex
bibliography: RJrefs.bib
---

@@

<!-- Tours -->
A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity.

@@

<!-- Manual tour paths -->
Ideally, the path would be intuitively user-generated from physical movement, be it through mouse or motion capture. Unfortunately, this type of dynamic control has proven difficult to capture for in R. Because of this, manual tours were not implemented in *tourr*. This research allows for the consumption, but not the generation, of such dynamic input. After the capture of an oblique user motion, the rotation needs to be applied to step 3 (rotation sequence) of the algorithm discussed below. In the section below we stick with a radial rotation where, $\theta$, the angle of in-projection-plane rotation is held constant.

<!-- Display -->
Spinifex utilizes two new animation packages, *plotly* [@sievert_plotly_2018] and *gganimate* [@pedersen_gganimate:_2019], to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

@@

<!-- remember mixing latex and rmarkdown bullets isn't trivial. -->
\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions.
  \item $\textbf{B} = (B_1,~ B_2)$, any of orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate.
  \item $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes.
  \item $\phi$, the angle of out-of-projection-plane rotation, coming into the manipulation space.
\end{itemize}

The algorithm primarily operates on the projection basis and utilizes the data only when making a display. The projection space can be viewed at any point in the process by pre-multiplying the data and plotting the first 2 variables.

@@

<!-- describe data. -->
The flea data, originally from @lubischew_use_1962, available in the R package *tourr* [@wickham_tourr_2011] is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

<!-- Projection basis -->
The data is defined. A basis set (ideally that views an interesting feature) should be provided to explore the sensitivity of the variables to the structure. To identify a projection containing an interesting feature, apply
a guided tour[@cook_interactive_2007] on the flea data. In a guided tour the projection sequence is selected by optimizing an index via hill-climbing. In this case, the holes index is selected. The holes index is maximized by when the projected observations are furthest from the center. Figure \@ref{fig:step0} shows a locally optimized projection for this data. The left plot displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projections. The right plot shows the projection of the data through the basis set described by the reference axes (left). Data points are colored and given point characters according to the species of the flea (the guided tour was unsupervised with this information).

(ref:step0-cap) Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables `aede2` and `tars1` contribute mostly orthogonal to the other variables. Select `aede2` as the manipulation variable to see how the structure of the projection changes as `aede2` is  rotated into and out of the projection.

@@

Call `view_basis()` on a basis to produce a *ggplot2* graphic similar to \@ref(fig:step0). Projection space is always available for display via the matrix multiplication $\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}$.

@@

<!-- select a manip var-->
In figure \@ref(fig:step0), above, the contributions of the variables `tars1` and `aede2` are mostly orthogonal to the contributions of the other four variables. These two variables explain the variation of the data between the purple and green species. We select `aede2` as the manip var, the variable to be manipulated as it typically has a larger contribution after the optimizing the holes index. The question that will be explored in the explanation of the algorithm is how important the variable `aede2` is to the separation of the clusters.

@@

<!-- Zero Vect, manip sp -->
Initialize a zero vector $e$ of $p$ elements. Because `aede2` is the fifth variable in the data, set the $k=5$-th element to one giving the manip var a full contribution in this dimension. Use the Gram-Schmidt process to orthonormalize the zero vector onto the basis yielding the 3D manipulation space, **M**.

\begin{align*}
  \textbf{e} &\leftarrow Orthonormalize_{GS}(\textbf{e}) w.r.t. the basis \\
  &= \textbf{e} - \langle \textbf{e},\textbf{B}_1 \rangle \textbf{B}_1 - \langle \textbf{e}, \textbf{B}_2 \rangle \textbf{B}_2 \\
  \\
  \textbf{M}_{[p,~3]} &= (\textbf{B}_1,\textbf{B}_2,\textbf{e})
\end{align*}

<!-- What the manip space facilitates -->
Adding this extra dimension to our basis plane allows for the coefficients of the specified variable to be changed. For example, the ability to lift a piece of paper, rather than being constrained to the motion on a table top. Orthonormalizing rescales the new depth vector while the projection down to 2D is the original basis, that is the first $d$ vectors remain constant. Imagine the reference axes (and projection plane) laying flat on a table, while a new dimension exists with axes projecting back onto the reference axes. An illustration of such can be seen below in figure \@ref(fig:step2). The manip var is highlighted, while the depths of the other variables are not depicted.

(ref:step2-cap) Manipulation space for controlling the contribution of `aede2` of standardized flea data. Basis selected by a holes-index guided tour. The Projection plane is shown in blue. The manipulation axis, in red, allows the coefficients of the manip var to be changed.

@@

The representation in \@ref(fig:step2) can be duplicated by calling the function `view_manip_space()`.

@@

<!-- illustration of axis manip -->
Imagine holding the red axis it is fixed to the origin. As it is manipulated the projection back onto the projection plane correspondingly moves. This is what happens in a manual tour. For a radial tour, fix $\theta$, the angle within the blue plane, and vary the sequence of $\phi$, the angle coming out of the projection plane. Conceptually, live manipulation on a 2D plane allows the user to dynamically control these angles, effectively changing the coefficients of the manip var, which then performs a constrained rotation on the remaining variables.

<!-- describe manip var path -->
For the demonstration of the radial tour, we define a sequence for $\phi$ that brings the initial contribution of the manip var to be maximized and then zeroed before returning to the initial position.

@@

Animations can be produced using the function `play_manual_tour()`. This function defaults to an HTML5 widget produced from *plotly*.

@@

The *tourr* package utilizes R's base graphics for the display of tours. *spinifex* allows tours to be used in rendered in *plotly* @sievert_plotly_2018 as an HTML5 object or *gganimate* @pedersen_gganimate:_2019 as .gif or .mp4 objects. Both of which build off *ggplot2* objects in internal functions. Sharing of animations is not trivial especially in print and static formats such as .pdf. Even with the use of computers and dynamic file formats capturing the correct resolution, aspect, and display is challenging and many formats quickly bloat file sizes. Keep in mind hosting options and exporting functions from *plotly*, *gganimate* and *tourr*.

@@

This article was created in *R* [@r_core_team_r:_2018], using *bookdown* [@xie_bookdown:_2016] and *rmarkdown* [@xie_r_2018], with code generating the examples inline. The source files for this article be found at [github.com/nspyrison/confirmation/](https://github.com/nspyrison/confirmation/). The source code for the *spinifex* package can be found at [github.com/nspyrison/spinifex/](https://github.com/nspyrison/spinifex/).

@@

<!-- summary and seague  -->
Tours, the dynamic linear projection of multivariate data, is an important aspect of data visualization extending the display of data-space as data dimensionality increases. This research has modified the algorithm producing manual tours, applied this functionality in *R* and offers extends the graphics offerings that can be used to display tours. The paragraphs below explore how this work might be extended.

@@

<!-- geometric and graphics display -->
The *tourr* package provides many other geometric displays with the `tourr::display_*()` family. These geometric options could be integrated into the *ggplot2* framework for display on *plotly* and *gganimate*. Additionally, the *animation* package @xie_animation:_2018 could be implemented for another graphics framework. However, *animation* builds from base graphs while *spinifex* utilizes *ggplot2* graphics.

@@

<!-- dynamic interaction/gui -->
Having a script only interaction with tours causes a significant barrier to entry. To a lesser extent, *plotly* offers some static interactions with the contained object, such as tooltips, brushing, and linking without communicating back to the R console. The development of a dynamic graphical user interface, perhaps with the use of a *shiny* [@chang_shiny:_2018] application, would mitigate the barrier to entry, allow for more rapid analysis, and offer an approachable demo tool. The user could easily switch between variables to control, adjust interpolation step angle, or flag/save specific frame basis sets.

>>>>

---
title: > 
  spinifex: manual control of dynamic linear projections of high-dimensional data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - School of Information Technology
    - Monash University, Melbourne, VIC 3800, Australia
    email:  Nicholas.Spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - School of Econometrics and Business Statistics
    - Monash University, Melbourne, VIC 3800, Australia
    email:  DiCook@monash.edu
abstract: >
  The set of dynamic linear projections of multivariate data collectively known as **tours** provide a unique dynamic visualization as the dimensionality of the data increase. Tours are particularly useful for understanding the structure held within multivariate data, and in association with techniques for dimension reduction, supervised, and unsupervised classification. The R package **tourr** offers a variety of path generators and geometric displays for conducting tours. This paper discusses an extension package, **spinifex**, that adds support for the path generation of manual tours and extends the display of tours to use with the contemporary animation packages, **plotly** and **gganimate**. Manual tours are used to explore the sensitivity of structure as the contributions of a manipulation variable are changed. This particularly useful after identifying a feature of interest.

  A recent paper {@wang_mapping_2018} visualizes the sensitivey of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolves more structrual information that is orthogonal to the original viewing plane.
keywords: manual tour, guided tour, grand tour, projection pursuit, high dimensional data, multivariate data, data visualization, statistical graphics, data science
output:
  bookdown::pdf_book:
    #base_format: rticles::rjournal_article #see sand box for hard code solution.
    fig_caption: yes
    toc: false
    includes:
      in_header: preamble.tex
bibliography: RJrefs.bib
---

@@

<!-- Tours -->
A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity. This is a crutial tool extending the dimensionality of vizualization which is large boon as datasets and models exist in increasing cardinality.

@@

<!-- Manual tour paths -->
Ideally, the path would be intuitively user-generated from physical movement, be it through mouse or motion capture. Unfortunately, this type of dynamic control has proven difficult to capture for in R. Because of this, manual tours were not implemented in **tourr** [@wickham_tourr_2011]. This research allows for the consumption, but not the generation, of such dynamic input. After the capture of an oblique user motion, the rotation needs to be applied to step 3 (rotation sequence) of the algorithm discussed below. In the section below we stick with a radial rotation where, $\theta$, the angle of in-projection-plane rotation is held constant.

<!-- Display -->
Spinifex utilizes two new animation packages, **plotly** [@sievert_plotly_2018] and **gganimate** [@pedersen_gganimate:_2019], to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

@@

<!-- remember mixing latex and rmarkdown bullets isn't trivial. -->
\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions.
  \item $\textbf{B} = (B_1,~ B_2)$, any orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate.
  \item $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes.
  \item $\phi$, the angle of out-of-projection-plane rotation, coming into the manipulation space.
\end{itemize}

The algorithm primarily operates on projection bases and utilizes the data only when making a display. The projection space can be viewed at any point in the process by pre-multiplying the data and plotting the first two variables.

@@

<!-- describe data. -->
The flea data, originally from @lubischew_use_1962, made available in **tourr**  is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

<!-- Projection basis -->
The data is defined. A basis set (ideally orienting to an interesting feature) must be provided as an initial orientation. One way to identify a projection containing interesting features is to apply a guided tour [@cook_interactive_2007]. In a guided tour the projection sequence is selected by optimizing an index fucntion via hill-climbing on the projection space. In this case, the holes index is selected and applied to standardized flea data. The holes index is maximized when the projected observations are furthest from the center. Figure \@ref{fig:step0} shows a locally optimized projection of the data. The left side displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projection. The right side shows the data as projectioned through the basis set described by the reference axes (left). Data points are colored and given shape according to the species (while the guided tour was unsupervised with this information).

(ref:step0-cap) Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables `aede2` and `tars1` contribute mostly orthogonal to the other variables.

@@

Call `view_basis()` on a basis to produce a **ggplot2** graphic similar to \@ref(fig:step0). Projection space is always available for display via the matrix multiplication $\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}$.

@@

<!-- select a manip var-->
In figure \@ref(fig:step0) the contributions of the variables `tars1` and `aede2` are mostly orthogonal to the contributions of the other four variables. These two variables explain the variation of the data distinguishing the purple group from the rest of the sample. Select `aede2` as the manip var, the variable to be manipulated, as it has a larger contribution to the projection. The question that will be explored is how important the variable `aede2` is to the separation of the clusters.

@@

<!-- Zero Vect, manip sp -->
Initialize a zero vector **e** of length $p$. Set the fifth element to on, as  `aede2` is the fifth variable in the data, giving the manip var a full contribution in this dimension. Use the Gram-Schmidt process to orthonormalize the zero vector onto the basis yielding the 3D manipulation space, **M**.

\begin{align*}
  \textbf{e} &\leftarrow Orthonormalize_{GS}(\textbf{e}) ~~~~~ w.r.t. the basis \\
  &= \textbf{e} - \langle \textbf{e},\textbf{B}_1 \rangle \textbf{B}_1 - \langle \textbf{e}, \textbf{B}_2 \rangle \textbf{B}_2 \\
  \\
  \textbf{M}_{[p,~3]} &= (\textbf{B}_1,\textbf{B}_2,\textbf{e})
\end{align*}

<!-- What the manip space facilitates -->
Adding this new dimension to our projection space allows for the coefficients of the manip var to be changed via rotation about the origin. For example, the ability to lift a piece of paper, rather than being constrained to a 2D table top. Orthonormalizing rescales the new vector while the projection down to 2D remains the original basis. Imagine the reference axes laying flat on a table, the a new dimension extends above the table, with axes projecting back onto the reference axes. An illustration of this is depicted in figure \@ref(fig:step2). The manip var is highlighted, while the depths of the other variables are not depicted.

(ref:step2-cap) Manipulation space for controlling the contribution of `aede2` of standardized flea data. Basis selected by a holes-index guided tour. The projection plane is shown in blue. The manipulation axis, in red, allows the coefficients of the manip var to be changed by rotation around the origin.

@@

The representation in figure \@ref(fig:step2) can be produced by calling the function `view_manip_space()`.

@@

<!-- illustration of axis manip -->
Imagine holding the red axis, one end fixed to the origin. As it is manipulated the projection back onto the projection plane correspondingly moves. This is what happens in a manual tour. For a radial tour, fix $\theta$, the angle within the blue plane, and vary the sequence of $\phi$, the angle coming out of the projection plane. Conceptually, live manipulation on a 2D plane allows the user to dynamically control these angles, effectively changing the coefficients of the manip var, which then performs a constrained rotation on the remaining variables.

<!-- describe manip var path -->
For the demonstration of the radial tour, a sequence is define for $\phi$ that brings the `XY` contributions of the manip var to a maximum and to zero before returning to the initial position.

@@

Animations can be produced using the function `play_manual_tour()`. This function defaults to an HTML5 widget produced from **plotly**.

@@

The **tourr** package utilizes R's base graphics for the display of tours. **spinifex** allows tours to be used in rendered in **plotly** @sievert_plotly_2018 as an HTML5 object or **gganimate** @pedersen_gganimate:_2019 as .gif or .mp4 objects. Both of which build off **ggplot2** objects in internal functions. Sharing of animations is not trivial especially in print and static formats such as .pdf. Even with the use of computers and dynamic file formats capturing the correct resolution, aspect, and display is challenging and many formats quickly bloat file sizes. Keep in mind hosting options and exporting functions from **plotly**, **gganimate** and **tourr**.

@@

This article was created in R [@r_core_team_r:_2018], using **bookdown** [@xie_bookdown:_2016] and **rmarkdown** [@xie_r_2018], with code generating the examples inline. The source files for this article be found at [github.com/nspyrison/confirmation/](https://github.com/nspyrison/confirmation/). The source code for the **spinifex** package can be found at [github.com/nspyrison/spinifex/](https://github.com/nspyrison/spinifex/).

@@

<!-- summary and seague  -->
Tours, the dynamic linear projection of multivariate data, is an important aspect of data visualization extending the display of data-space as data dimensionality increases. This research has modified the algorithm producing manual tours, applied this functionality in R and offers extends the graphics offerings that can be used to display tours. The paragraphs below explore how this work might be extended.

@@

<!-- geometric and graphics display -->
The **tourr** package provides many other geometric displays with the `tourr::display_*()` family. These geometric options could be integrated into the **ggplot2** framework for display on **plotly** and **gganimate**. Additionally, the **animation** package @xie_animation:_2018 could be implemented for another graphics framework. However, **animation** builds from base graphs while **spinifex** utilizes **ggplot2** graphics.

@@

<!-- dynamic interaction/gui -->
Having a script only interaction with tours causes a significant barrier to entry. To a lesser extent, **plotly** offers some static interactions with the contained object, such as tooltips, brushing, and linking without communicating back to the R console. The development of a dynamic graphical user interface, perhaps with the use of a **shiny** [@chang_shiny:_2018] application, would mitigate the barrier to entry, allow for more rapid analysis, and offer an approachable demo tool. The user could easily switch between variables to control, adjust interpolation step angle, or flag/save specific frame basis sets.
